6.878 Final Report

### 2013-12-10 report/paper.tex
changes to the report.

  Of the two time course datasets we tried this approach, only one of them shows a corresponding enrichment of biological terms associated identified clusters. The root cause for the failure in the other dataset is yet to be identified.

@@

\section{Background}

@@

  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering). We only make a brief mention of visualization and annotation for completion.

@@

The RNASeq dataset posted on GEO is the output of topHat (~120GB) and we performed the following preprocessing.
\begin{description}
 \item a. Cufflinks analysis to derive gene expression abundance in FPKMs
 \item b. Filtering and Quantile normalization of the cufflinks output.
 \item c. We then created multiple views of this expression data to get a feel for how the
		data looks.
\end{description}

In addition to the above we tested our methodology on the following two time course datasets.

@@

Given the distance matrix generated by alignment, we were able to cluster similar sequences.
Our clustering was based on the alignment scores generated above as distance.
One question that needed to be resolved was how determine the optimal number of
clusters.We have tried the following two approaches towards this
a. Pairwise clustering until we reach a threshold of the alignment score between
two clusters.
b. Affinity propagation based clustering to automatically determine the number of
clusters.The pairwise heirarchical clustering with an emperical value of threshold based
out of experiments gives us fewer (about half) the clusters than the affinity propogation ,
however the clusters generated by affinity propogation appear to be more cohesive

Our first test is to check how the alignment and clustering compared to the
preferred UPGMA clustering. We generated a heirarchical clustering output on the
same preprocessed dataset with the help of an external tool ‘Cluster 3.0’
developed by Michael Eisen at stanford.
Shown below are three heatmaps the first two heatmaps are from clusters
generated by our implementation of alignment and clustering (affinity prop
followed by heirarchical) the third generated by cluster 3.0. In purple is an
example of a cluster that based on the signal appears to belong together but is
spread out in multiple clusters if alignment is not considered for clustering.

@@

\subsection{Enrichment Analysis}
After identifying clusters, we wanted a way to efficiently describe each cluster, as well as to measure the effectiveness of our alignment and clustering algorithm.
To analyze the efficacy of alignment in clustering, we used Fischer's Exact test for Gene Ontology annotations.
Gerald Quon provided us with an \verb!R! script and Gene Ontology data for annotation testing of our clusters.
While, visually, the clusters showed alignment, the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure. However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means (Fig. \ref({fig:enrichment}). It shows that, to some degree, our alignment and clustering technique performs better than without alignment.\\

@@

\subsection{Visualization}
In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3! that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID! gene groups by cluster (Fig. \ref({fig:enrichment}).
\section{Future Goals}
We would re-do the pre-processing step or attempt to get the pre-processed dataset before reanalyzing our data.

@@

\section{Commentary on peer-review process}
Iris: I thought that it was good to get feedback, but at the same time, our proposal was missing a lot of detail on purpose because we were not sure at the time exactly on what we were going to do, and a lot of the feedback was on lack of detail, which was not that helpful. It was nice to see other people's ideas.

@@

Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.

>>>>

  Of the three time course datasets we tried this approach, only two of them appear to show a corresponding enrichment of biological terms associated with the identified clusters. The root cause for the failure for clusters to show significant enrichment in the third dataset is yet to be identified.

@@

\section{Background}
	\begin{figure}[H]
\centering
\includegraphics[scale=0.65]{Introduction.png}
\caption{An individual patients molecular profile captured across time points.}
\label{fig:PatientTimeCourses}
\end{figure}

@@

  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering). We make a brief mention of visualization and annotation for completion.

@@

The RNASeq dataset posted on GEO is the output of topHat (~120GB) and we performed the following preprocessing.
\begin{description}
 \item a. Cufflinks analysis to derive gene expression abundance in FPKMs
 \item b. Filtering,Quantile normalization,vector normalization of the cufflinks output.
 \item c. We then created multiple views of this expression data to get a feel for how the
		data looks.
\end{description}

In addition to the above we tested our methodology on the following two time course gene expression datasets available on GEO. No additional preprocessing was done to these datasets.

@@

Given the distance matrix generated by alignment, we were able to cluster similar sequences.
Our clustering was based on the alignment scores generated above as distance.
One question that needed to be resolved was how determine the optimal number of
clusters.We have tried the following two approaches towards this
a. Pairwise clustering until we reach a threshold of the alignment score between
two clusters.
b. Affinity propagation based clustering to automatically determine the number of
clusters.The pairwise heirarchical clustering with an emperical value of threshold based
out of experiments gives us fewer clusters than the affinity propogation . However the clusters 
generated by affinity propogation appear to be more cohesive

We compared our clustering and alignment to the standard UPGMA clustering. We generated a heirarchical clustering output on the same preprocessed dataset with the help of an external tool ‘Cluster 3.0’ developed by Michael Eisen at stanford.
Shown below are three heatmaps the first two heatmaps are from clusters generated by our implementation of alignment and clustering (affinity prop followed by heirarchical) the third generated by cluster 3.0. In yellow is an example of a cluster that based on the signal appears to belong together but is spread out in multiple clusters if alignment is not considered for clustering.

@@

\subsection{Enrichment Analysis}
After identifying clusters, we wanted a way to efficiently describe each cluster, as well as to measure the effectiveness of our alignment and clustering algorithm.

To analyze the efficacy of alignment in clustering, we used multiple approaches 1. Fischer's Exact test for Gene Ontology annotations. 2. We ran the gene list generated by our clusters through DAVID functional annotation tool.
Gerald Quon provided us with an \verb!R! script and Gene Ontology data for annotation testing of our clusters.

While, visually, the clusters showed alignment for GSE33029 dataset , the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure. However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means (Fig. \ref({fig:enrichment}). It shows that, to some degree, our alignment and clustering technique performs better than without alignment.\\

@@

We have been unable to pinpoint the cause for the lack of enrichment annotations for our clusters on the GSE33029 dataset.We suspected that we are missing a noise correction step for the GSE33029 dataset that may be resulting in no enrichment showing up and reanalyzed with additional normalizations in no vain. We suspect the authors may have not fully specified the steps they took to preprocessing this dataset and we may be missing some correction step.

To check if our approach works we decided to rerun our analysis on two additional and independent timecourse datasets from GEO (GSE19392 GSE675) . With clusters generated from these datasets we do significant enrichment numbers showing up for pathways and other annotation terms. The KEGG pathway results for a few of our clusters are as shown below.

\begin{figure}[H]
\centering
\includegraphics[scale=0.65]{3Clusters_PathwayCombined.png}
\caption{KEGG Pathway analysis on 3 clusters from GSE19392 dataset using david.}
\label{fig:KEGGPathway.}
\end{figure}
\subsection{Visualization}

In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3! that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID! gene groups by cluster (Fig. \ref({fig:enrichment}).
\section{Future Goals}
We would re-do the pre-processing step or attempt to get the pre-processed dataset before reanalyzing our data.

@@

Dalesh: Working on the project was an enjoyable experience , a few highlights are as follows
a. Opportunity to work in a team and have a mentor to guide us.
b. Applying our learning on a couple of real datasets gave additional insights.
c. The two interim feedback checkpoints

\section{Commentary on peer-review process}
Iris: I thought that it was good to get feedback, but at the same time, our proposal was missing a lot of detail on purpose because we were not sure at the time exactly on what we were going to do, and a lot of the feedback was on lack of detail, which was not that helpful. It was nice to see other people's ideas.
Dalesh: The peer review process was fun . And i agree with Iris's comment on the feedback we received , i recollect giving ratings on the table without having completely understood what the authors were trying to communicate.

@@

Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.
Dalesh: Most of the strategy and design was arrived at by brainstorming together , my specific implementation contributions were in data preprocessing on the amazon EC2 instances and clustering , I was responsible for processing and analyzing the GSE19392 dataset , we split the analysis of the GSE33029 dataset . Iris also did the implementations for the user interface.
